Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2014-09-17 paper/paper.tex
minor edit suggested by Jeff Tested everything on bennet.
edits to introduction in paper Tested everything on crystal.

\section{Introduction}
\fixme{Huang and those who reference him use the terms 'numerical
  deterministic reaction-diffusion model' and 'stochastic version of
  this model' for the stochastic.} It is vital that during the process
of bacterial cell division a cell avoid minicelling, or splitting into
daughter cells with lopsided volumes.  Instrumental to this process in
\emph{Escherichia coli} is a long FtsZ polymer chain that develops on
the cell wall in the center region of the cell, helping dictate the
plane of division~\cite{adams2009bacterial,
  lutkenhaus2007assembly}. Previous experimental studies have shown
that the MinC protein, known to inhibit the FtZ
polymer\cite{shen2010examination}, exhibits pole to pole oscillatory
behavior in conjunction with the MinD protein and a MinE that tends to
be more localized in the center of the cell~\cite{hu1999topological,
  fu2001mine, shapiro2009and, yu1999ftsz, raskin1999rapid}. The MinC
protein will then have a higher time averaged concentration in the
cell poles, as opposed to the center region of the cell, aiding in
prohibiting the FtZ from developing in the wrong region.

Studies have shown that the cell division process is capable of
preventing minicelling for fairly perturbed cell
shapes\cite{touhami2006temperature}. Varma et al. studied one
particular example of a perturbed shape, a three-pronged tube, and
found both experimental and simulation results\cite{varma2008min} that
show oscillation. These oscillations seem to seek out the extreme
poles in the cell shape\cite{corbin2002exploring}
\cite{juarez2010changes}, and effectively prevent minicelling in most
cases. However, Mannik et al. have also shown that there are
limitations to this mechanism\cite{mannik2010bacteria}
\cite{mannik2009bacterial}. One instance of this behavior is
observable when the cell has been forced into a flattened, irregular
shape. Oscillations in this sort of cell shape are far from the
regular sorts of oscillations seen in the computational work done by
Huang et al\cite{huang2003dynamic}, who simulate more regular
shapes. Thus there is an opportunity here to test previously used
simulation models against more extreme experimental cases than have
been seen thus far.  In this paper we apply a deterministic method and
a stochastic method of simulation and compare them against Mannik's
experimental findings.

A significant amount of work has been done to develop protein reaction
and diffusion models that exhibit accurate macroscopic dynamics of the
MinD protein system. Early models involve free proteins that affect
each others' rates of diffusion and membrane attachment, but do not
combine into compound states~\cite{meinhardt2001pattern}.  In 2003
Huang improved upon this work with a simple and very successful
simulation model based on MinD-MinE combination, ATPase hydrolysis,
and MinD membrane attachment that exhibits accurate MinD oscillations
in cylindrical cells~\cite{huang2003dynamic}. In this model
cytoplasmic MinD is more likely to attach to the membrane when MinD is
already clustered there (following observed non-linear attachement of
minD on the cell membrane), and is stationary once attached.  A number
of studies have used an approach similar in that they do not rely on
the ability of MinD to move along the walls and
cluster~\cite{kruse2007experimentalist, meinhardt2001pattern,
  drew2005polymerization, fange2006noise, kerr2006division}, while
studies have been made as well of models which rely on MinD mobility
and attraction on the cell membrane~\cite{kruse2002dynamic,
  howard2005cellular}.  Experimental studies of the Min system's
association with the cell membrane have also been
made~\cite{hsieh2010direct}\cite{mileykovskaya2003effects}.

Variations of the Huang 2003 model that stochasticly account for
variations of molecular interaction \cite{fange2006noise} and
monte-carlo simulations that implement this stochastic version of
Huang's mean field reaction rates confirm the major results obtained
by Huang's model, and more successfully predict experimentally
observed oscillations in round cell
phenotypes~\cite{drew2005polymerization, fange2006noise,
  huang2004min}. Biochemical models of broader scope have also been
used to study the MinD system and show consistent
results.\cite{arjunan2010new}.  However, in general, the results of
the stochastic and monte-carlo simulations have so far yeilded similar
results to those given by Huang's original, deterministic method of
simulation.\fixme{Look at the papers again to verify and this and also
  figure out and cite who says that the stochastic and the
  deterministic methods of simulation have given similar results.
  Saying more about these discussions is needed here, I think.
  Something to the effect that until now there's no compelling reason
  to avoid the deterministic simulations.}

>>>>

\section{Introduction}
\fixme{Huang and those who reference him use the terms 'numerical
  deterministic reaction-diffusion model' and 'stochastic version of
  this model' for the stochastic.} It is vital that during the process
of bacterial cell division a cell avoid minicelling, or splitting into
daughter cells with lopsided volumes.  Instrumental to this process in
\emph{Escherichia coli} is a long FtsZ polymer chain that develops on
the cell wall in the center region of the cell, helping dictate the
plane of
division~\cite{adams2009bacterial,lutkenhaus2007assembly}. Previous
experimental studies have shown that the MinC protein, known to
inhibit the FtZ polymer\cite{shen2010examination}, exhibits pole to
pole oscillatory behavior between both ends of the wild-type
pill-shaped cell.  It thus has a higher time averaged concentration in
the cell poles than in the center region, which aides in prohibiting
the FtZ from developing in the wrong region.  The MinC is recruited to
these poles by MinD, which ineracts with another protein, MinE, that
is neccesary for MinD oscillation.~\cite{hu1999topological,
  fu2001mine, shapiro2009and, yu1999ftsz, raskin1999rapid}.

Previous studies have shown that the minD protein system is capable of
exhibiting oscillations in round shapes as well as in connected three
pronged tube shapes~\cite{varma2008min}, in which the oscillations
seem to seek out the extreme poles in the
cell~\cite{corbin2002exploring} \cite{juarez2010changes}. However,
Mannik \emph{et al.} have also shown that there are limitations to
this capability~\cite{mannik2010bacteria}
\cite{mannik2009bacterial}. They have experimentally forced E. Coli
cells into flattened irregular shapes of $0.25\mu m$
thickness. Oscillations in this type of cell shape lose their
regularity, both temporally and spatially. This allows for an
opportunity to test previously studied MinD simulation models against
more extreme experimental cases than have been seen thus far.

A significant amount of work has been done to develop protein reaction
and diffusion models that exhibit accurate macroscopic dynamics of the
MinD protein system. Early models involve free proteins that affect
each others' rates of diffusion and membrane attachment, but do not
combine into compound states~\cite{meinhardt2001pattern}.  In 2003
Huang improved upon this work with a simple and very successful
simulation model based on MinD-MinE combination, ATPase hydrolysis,
and MinD membrane attachment that exhibits accurate MinD oscillations
in cylindrical cells~\cite{huang2003dynamic}. In this model
cytoplasmic MinD is more likely to attach to the membrane when MinD is
already clustered there (following observed non-linear attachement of
minD on the cell membrane), and is stationary once attached.  A number
of studies have used an approach similar in that they do not rely on
the ability of MinD to move along the walls and
cluster~\cite{kruse2007experimentalist, meinhardt2001pattern,
  drew2005polymerization, fange2006noise, kerr2006division}, while
studies have been made as well of models which rely on MinD mobility
and attraction on the cell membrane~\cite{kruse2002dynamic,
  howard2005cellular}.  Experimental studies of the Min system's
association with the cell membrane have also been
made~\cite{hsieh2010direct,mileykovskaya2003effects}.

Variations of the Huang 2003 model that stochasticly account for
variations of molecular
interaction~\cite{kerr2006division,fange2006noise} and monte-carlo
simulations that implement this stochastic version of Huang's mean
field reaction rates confirm the major results obtained by Huang's
model, and more successfully predict experimentally observed
oscillations in round cell phenotypes~\cite{fange2006noise,
  huang2004min}. The successful simulations of three pronged cells
mentioned above are stochastic
simulations~\cite{varma2008min}. Biochemical models of broader scope
have also been used to study the MinD system and show consistent
results~\cite{arjunan2010new}.

\fixme{There are not any direct comparisons of the deterministic and
  stochastic methods that I can find other than the fange2006 paper
  cited above for the round cells.  Those showed that the
  deterministic couldn't predict oscillations in perfectly round
  cells, but could in almost round ellipses.  It seems that once the
  stochastic methods began to be used, nobody really used the
  deterministic method any longer.  Not sure what to say about this in
  order to introduce what we do here properly.}

In this paper we apply a deterministic method and a stochastic method
of simulation and compare them against Mannik's experimental findings.
